Large-scale hybrid ad hoc network for mobile platforms: Challenges and Experiences

### 2019-02-11 paper.tex
All done except experimental results

\section{Introduction}
Mobile devices with programmable platforms such as android and iOS
have steadily grown over the last decade, surpassing the 2 billion
mark \footnote{https://www.statista.com/statistics/330695/number-of-smartphone-users-worldwide/}. MANETs
have been widely studied given their decentralized nature and
potential for new applications
\cite{loo-manet-2011,perkins-ad-hoc-2001}. Most of the prior work on
p2p networks has focused on analytical and simulation-based study of
MANET behavior
\cite{zhang-topology-manet-2015,marti-misbehavior-manet-2000,mauve-pos-routing-manet-2001}. Given
the outstanding practical challenges in deployoing a large number of
physical nodes, real-world implementations have been limited and have
not reached mass scale \cite{kiess-manet-impl-2007}. However, with the
growth of smartphones, large-scale real-world implementations may
become feasible. This paper describes a real-world implementation of a
p2p delay-tolerant network, called Mesh Network Alerts (MNA), for
relaying severe weather information to millions of mobile device users
as part of the Weather Channel
app\footnote{https://weather.com/apps/ibm/meshnetworkalerts} on both
android and iOS platforms.

@@

The following describes how the c2p2p architecture addresses the
challenges identified above along with known limitations. Further, to
realize this architecture on mobile devices, we evaluated a variety of
protocols on both Android and iOS. As described later, WiFi DNS on
Android, Bluetooth Nearby on Android, and Blutooth low energy (BTLE)
on both Android and iOS are the only protocols that were found
effective. Actual techniques for discovery, advertisement, and
connection vary across these protocols and are described next.
%
\subsection{Operating system restrictions}
\label{sec:os}
MNA works around the background activity and user permission
restrictions via innovative techniques, without resorting to hacks
that may violate user security or App-store guidelines.
Tables~\ref{tab:android} and~\ref{tab:ios} summarize the various
protocols we have experimented with using WiFi and Bluetooth
interfaces, along with key findings as described here.  The protocols
actually deployed are in bold.
%
%
\begin{table}[H]
\caption{Android protocols and key findings}
\label{tab:android}
\centering
\begin{tabular}{|c|l|r|r|r|c|l|}
\hline
& \multicolumn{1}{c|}{Protocol} & \multicolumn{1}{c|}{Throughput} & \multicolumn{1}{c|}{Battery} & \multicolumn{1}{c|}{Range} & \multicolumn{1}{c|}{iOS} & \multicolumn{1}{c|}{Issues}\\
&                               & \multicolumn{1}{c|}{Avg. kbps}  & \multicolumn{1}{c|}{per hour} & \multicolumn{1}{c|}{LoS ft} &                  &                            \\
\hline
\parbox[t]{2mm}{\multirow{3}{*}{\rotatebox[origin=c]{90}{WiFi}}} & \textbf{DNS} & 0.2 & 2\% & 600 & No & \\
& WiDi & 2000 & 3\% & 600 & No & Needs DNS\\
& Hotspot & 2000 & 5\% & 600 & No & Permission\\
\hline
\parbox[t]{2mm}{\multirow{3}{*}{\rotatebox[origin=c]{90}{BT}}} & \textbf{Classic} & 50 & 2\% & 600 & No & \\
& \textbf{Nearby} & 50 & 2\% & 600 & No & \\
& \textbf{BTLE} & 50 & 2\% & 600 & Yes & \\
\hline
\end{tabular}
\end{table}
%
%
\begin{table}[H]
\caption{iOS protocols and key findings}
\label{tab:ios}
\centering
\begin{tabular}{|c|l|r|r|r|c|l|}
\hline
& \multicolumn{1}{c|}{Protocol} & \multicolumn{1}{c|}{Throughput} & \multicolumn{1}{c|}{Battery} & \multicolumn{1}{c|}{Range} & \multicolumn{1}{c|}{Android} & \multicolumn{1}{c|}{Issues}\\
&                               & \multicolumn{1}{c|}{Avg. kbps}  & \multicolumn{1}{c|}{per hour} & \multicolumn{1}{c|}{LoS ft} &                  &                            \\
\hline
\parbox[t]{2mm}{\multirow{2}{*}{\rotatebox[origin=c]{90}{WiFi}}} & Bonjour & 0.2 & 5\% & 200 & No & Battery\\
& MPC & 2000 & 5\% & 200 & No & Battery\\
\hline
& \textbf{BTLE} & 50 & 1\% & 800 & Yes & \\
\hline
\end{tabular}
\end{table}
%

@@

%
\subsection{Power constraints}
\label{sec:power}
%
Due to indefinite forground services and continuous discovery and
advertisement, this is primarily an issue on Android. Our approach
here is two-pronged. Firstly, via extensive experiments on a large
number of devices, we fine-tune the algorithms governing the intervals
at which discovery and advertisements occur. In a nutshell, receiving
new information during a period causes MNA to be more aggressive in
discovery and advertisement. Similarly, lack of new information for a
period makes the device less aggressive. Secondly, we allow ``wake
up'' messages to be broadcast in the network ahead of an anticipated
severe weather event. When devices receive such messages, they
schedule themselves to remain aggressive during the specified window
of time. Outside of this window, the device can afford to have long
sleep cycles and conserve power. With these techniques, our testing
shows less than 2\% battery consumption per hour on most device
models.
%
\subsection{Testing p2p networks}
\label{sec:texting}
%
We developed test automation tools and processes such that multiple
devices can be controlled from a single test station and follow
prescribed steps to generate, send, and receives messages to play out
a test scenario. Finally, the framework allows automated analysis of
the observations to determine the test result. This capability was
instrumental in uncovering bugs at a fast pace to meet the aggressive
timeline and avoid the cost of acquisition. Further, the automation
framework is general and can be expressly applied to test other apps
in this fashion.

@@

%
\subsection{Trust in information}
\label{sec:trust}
%
In our approach, since the origin of weather information is the
Weather Channel service, digital signatures can be attached to all
weather alerts broadcast from the service. The mobile application is
distributed with the corresponding public key so that digital
signatures from the service can be verified. If a message fails such a
verification, it is discarded and not forwarded any further.
%
\section{Algorithms}
\label{sec:algo}
%
\subsection{WiFi DNS}
\label{sec:wifi}
%
DNS service discovery is a widely supported protocol
\cite{cheshire-dns-sd-2013}. On Android, WiFi DNS improvises on this
standard such that the exchange of information happens without making
network connections at all. This is achieved by splitting the message
payloads into small enough chunks and stuffing them into service
advertisements as txt records and broadcast over multiple
advertisements in a quick succession. Since there are no connections
being made, the messages can be exchanged without prompting the user
to grant permissions for each connection. Since no peer connections
are needed and advertisements and discovery are broadcast, WiFi DNS is
a broadcast protocol. On iOS, Bonjour protocol is built on DNS service
discovery but is power-hungry and does not interoperate with WiFi DNS
on Android. Algorithm~\ref{algo:dns} outlines the peer process for
WiFi DNS. Power conservation via time intervals is achieved by
manupulating $T_{ad}$ and $T_{disc}$ based on traffic.

@@

\begin{algorithm}[htb!]
\label{algo:btle}
\DontPrintSemicolon % Some LaTeX compilers require you to use \dontprintsemicolon instead
\SetAlgoLined
\KwIn{Peer id $p_{id}$, App id $a_{id}$, Max peripherals $max_p$}
 $S_{name} \gets$ \textsf{'MNA'}\;
 $\mathbb{M} \gets \varnothing$\;
 $C \gets$ new Central() in new thread\;
 $P \gets$ new Peripheral() in new thread\;
 $T \gets$ new data sender thread\;
 $C$.bleDiscovery($p_{id}$, $S_{name}$)\;
 $P$.bleAdvert($p_{id}$, $S_{name}$)\;
 $P_{con} \gets 0$\;

>>>>

\section{Introduction}
Mobile devices with programmable platforms such as android and iOS
have steadily grown over the last decade, surpassing the 2 billion
mark \footnote{https://www.statista.com/statistics/330695/number-of-smartphone-users-worldwide/}. MANETs
have been widely studied given their decentralized nature and
potential for new applications
\cite{loo-manet-2011,perkins-ad-hoc-2001}. Most of the prior work on
p2p networks has focused on analytical and simulation-based study of
MANET behavior
\cite{zhang-topology-manet-2015,marti-misbehavior-manet-2000,mauve-pos-routing-manet-2001}. Given
the outstanding practical challenges in deployoing a large number of
physical nodes, real-world implementations have been limited and have
not reached mass scale \cite{kiess-manet-impl-2007}. However, with the
growth of smartphones, large-scale real-world implementations may
become feasible. This paper describes a real-world implementation of a
p2p delay-tolerant network, called Mesh Network Alerts (MNA), for
relaying severe weather information to millions of mobile device users
as part of the Weather Channel app \cite{mna} on both android and iOS
platforms.

@@

The following assesses the c2p2p architecture relative to the
challenges identified above along with known limitations. Further, to
realize this architecture on mobile devices, we evaluated a variety of
protocols on both Android and iOS. As described later, WiFi DNS on
Android, Bluetooth Nearby on Android, and Blutooth low energy (BTLE)
on both Android and iOS are the only protocols that were found
effective. Actual techniques for discovery, advertisement, and
connection vary across these protocols and are described next.
%
\subsection{Operating system restrictions}
\label{sec:os}
MNA works around the background activity and user permission
restrictions via innovative techniques, without resorting to hacks
that may violate user security or App-store guidelines.
Tables~\ref{tab:android} and~\ref{tab:ios} summarize the various
protocols we have experimented with using WiFi and Bluetooth
interfaces, along with key findings as described here.  The protocols
actually deployed are in bold.
%
%
\begin{table}[H]
\caption{Android protocols and key findings}
\label{tab:android}
\centering
\begin{tabular}{|c|l|r|r|r|c|l|}
\hline
& \multicolumn{1}{c|}{Protocol} & \multicolumn{1}{c|}{Throughput} & \multicolumn{1}{c|}{Battery} & \multicolumn{1}{c|}{Range} & \multicolumn{1}{c|}{iOS} & \multicolumn{1}{c|}{Issues}\\
&                               & \multicolumn{1}{c|}{Avg. kbps}  & \multicolumn{1}{c|}{per hour} & \multicolumn{1}{c|}{LoS ft} &                  &                            \\
\hline
\parbox[t]{2mm}{\multirow{3}{*}{\rotatebox[origin=c]{90}{WiFi}}} & \textbf{DNS} & 0.2 & 2\% & 600 & No & \\
& WiDi & 2000 & 3\% & 600 & No & Needs DNS\\
& Hotspot & 2000 & 5\% & 600 & No & Permission\\
\hline
\parbox[t]{2mm}{\multirow{3}{*}{\rotatebox[origin=c]{90}{BT}}} & \textbf{Classic} & 50 & 2\% & 600 & No & \\
& \textbf{Nearby} & 50 & 2\% & 600 & No & \\
& \textbf{BTLE} & 50 & 2\% & 600 & Yes & \\
\hline
\end{tabular}
\end{table}
%
%
\begin{table}[H]
\caption{iOS protocols and key findings}
\label{tab:ios}
\centering
\begin{tabular}{|c|l|r|r|r|c|l|}
\hline
& \multicolumn{1}{c|}{Protocol} & \multicolumn{1}{c|}{Throughput} & \multicolumn{1}{c|}{Battery} & \multicolumn{1}{c|}{Range} & \multicolumn{1}{c|}{Android} & \multicolumn{1}{c|}{Issues}\\
&                               & \multicolumn{1}{c|}{Avg. kbps}  & \multicolumn{1}{c|}{per hour} & \multicolumn{1}{c|}{LoS ft} &                  &                            \\
\hline
\parbox[t]{2mm}{\multirow{2}{*}{\rotatebox[origin=c]{90}{WiFi}}} & Bonjour & 0.2 & 5\% & 200 & No & Battery\\
& MPC & 2000 & 5\% & 200 & No & Battery\\
\hline
& \textbf{BTLE} & 50 & 1\% & 800 & Yes & \\
\hline
\end{tabular}
\end{table}
%

@@

%
\subsection{Power constraints}
\label{sec:power}
%
Due to indefinite forground services and continuous discovery and
advertisement, this is primarily an issue on Android. Our approach
here is two-pronged. Firstly, via extensive experiments on a large
number of devices, we fine-tune the algorithms governing the intervals
at which discovery and advertisements occur. In a nutshell, receiving
new information during a period causes MNA to be more aggressive in
discovery and advertisement. Similarly, lack of new information for a
period makes the device less aggressive. Secondly, we allow the
centrasl service to broadcast``wake up'' messages ahead of an
anticipated severe weather event.  When devices receive such messages,
they schedule themselves to remain aggressive during the specified
window of time.  Outside of this window, the device can afford to have
long sleep cycles and conserve power. With these techniques, our
testing shows less than 2\% battery consumption per hour on most
device models.
%
\subsection{Testing p2p networks}
\label{sec:texting}
%
We developed novel test automation tools and processes that control
multiple devices from a single test station and follow prescribed
steps to generate, send, and receives messages to play out a test
scenario.  The framework allows automated analysis of the observations
to determine the test result. This capability was instrumental in
uncovering bugs at a fast pace with tens of physical devices emloyed
for automated testing.  Further, the automation framework is general
and can be expressly applied to test other apps in this fashion.

Figure~\ref{fig:test_arch} shows the overall test automnation
architecture. A tester specifies high-level test cases and expected
results in a platform and device agnostic fashion. Mobile devices run
a UI proxy for executing platform and device-specific UI command while
a central laptop terminal acts as the overall orchestrator and
translates high-level test cases and distributes them to
devices. Since the orchestrator needs a wired access to the devices
and the devices cannot be programmed to move (unlike drones and bots),
testing of range and mobility patterns needs to done manually, which
is a known limitation.

A key innovation here is the platform-independent language for
specifying high-level test cases and a small set of primitive commands
it translates to. We describe here the list of primitive commands and
a screen recording across seven Android devices demonstrates this
powerful capability \cite{mna-test-demo}.
\begin{itemize}
\item \textsf{syncAll} Wait for all devices to reach this point
\item \textsf{object.find} Find a UI object by name or position
\item \textsf{object.click} Click a UI object
\item \textsf{sleep} Wait for the specified duration
\item \textsf{setupSystemAlert} Respond to a system popup
\item \textsf{text.set} Highlight a text input box
\item \textsf{text.write} Write text to the text box
\item \textsf{swipe} Swipe to the left/right or up/down
\end{itemize}

@@

%
\subsection{Trust in information}
\label{sec:trust}
%
In c2p2p, since the origin of information is always the central
service, digital signatures can be attached to all messages
originating from the service. The mobile application is distributed
with the corresponding public key so that digital signatures from the
service can be verified.  If a message fails such a verification, it
is discarded and not forwarded any further.

Peers cannot originate messages however as then a malicious peer can
join the network, exchange keys with other peers, start generating,
signing, and spreading fake information. This has been a longstanding
problem for open decentralized systems. Requiring all peers to
register their identity with a certificate authority and reporting the
originators of fake information may discourage malicious behavior.
Clearly, this is even harder to solve than the spread of fake
information over the Internet because ground truth is hard to find and
malicious peers can report genuine sources of information as well.
%
\section{Algorithms}
\label{sec:algo}
%
\subsection{WiFi DNS}
\label{sec:wifi}
%
DNS service discovery is a widely supported protocol
\cite{cheshire-dns-sd-2013}. On Android, WiFi DNS improvises on this
standard such that the exchange of information happens without making
network connections at all. This is achieved by splitting the message
payloads into small enough chunks and stuffing them into service
advertisements as txt records and broadcast over multiple
advertisements in a quick succession. Since there are no connections
being made, the messages can be exchanged without prompting the user
to grant permissions for each connection. Since no peer connections
are needed and advertisements and discovery are broadcast, WiFi DNS is
a broadcast protocol. On iOS, Bonjour protocol is built on DNS service
discovery but is power-hungry and does not interoperate with WiFi DNS
on Android. Algorithm~\ref{algo:dns} outlines the peer process for
WiFi DNS. Power conservation via time intervals is achieved by
manupulating $T_{ad}$ and $T_{disc}$ based on traffic.

@@

Algorithm~\ref{algo:btle} outlines the peer process for BTLE. The
biggest difference with WiFi DNS is that discovery and advertisements
happen only once per app launch. Secondly, the entire peer process is
event-driven, so the app can sleep when there are no events to
process. The list of forwarders for each message $M.fwdList$ is key in
controlling the duplicates and in turn network utilization without
overhead messages. In practice, limiting number of Peripherals a
Central connects to enables reliable communications, specified by
$max_p$.

\begin{algorithm}[htb!]
\label{algo:btle}
\DontPrintSemicolon % Some LaTeX compilers require you to use \dontprintsemicolon instead
\SetAlgoLined
\KwIn{Peer id $p_{id}$, App id $a_{id}$, Max peripherals $max_p$}
 $S_{name} \gets$ \textsf{'MNA'}\;
 $\mathbb{M} \gets \varnothing$\;
 $C \gets$ new Central() in new dispatch queue\;
 $P \gets$ new Peripheral() in new dispatch queue\;
 $T \gets$ new data sender dispatch queue\;
 $C$.bleDiscovery($p_{id}$, $S_{name}$)\;
 $P$.bleAdvert($p_{id}$, $S_{name}$)\;
 $P_{con} \gets 0$\;
